---
title: "HEART DISEASE PREDICTION ANALYSIS"
author: "Anindita Bag"
output: html_document
---

## STATISTICAL ANALYSIS USING R PROGRAMMING

-   [**LIBRARIES REQUIRED FOR THE ANALYSIS**:]{.underline}

```{r }
library(tidyverse)
library(ggplot2)
library(dplyr)
library(psych)
```

-   [**DATA**]{.underline}

    ***Getting the Data Into R:***

```{r}
data=read.csv("C:/Users/STUDENT/Downloads/anindita.csv")
```

The first part of the analysis is to read the data set and clean the column names up a bit-

***Prepare column names:***

```{r}
names=c("Age",
           "Sex",
           "Chest_Pain_Type",
           "Resting_Blood_Pressure",
           "Serum_Cholesterol",
           "Fasting_Blood_Sugar",
           "Resting_ECG",
           "Max_Heart_Rate_Achieved",
           "Exercise_Induced_Angina",
           "ST_Depression_Exercise",
           "Peak_Exercise_ST_Segment",
           "Num_Major_Vessels_Flouro",
           "Thalassemia",
           "Diagnosis_Heart_Disease")
```

***Apply Column names to the dataframe***

```{r}
colnames(data)=names
```

***Glimpse data to verify new column names are in place***

```{r}
data%>%glimpse()
```

***Display the first few rows in the data***

```{r}
head(data)
```

-   [**EXPLORATORY DATA ANALYSIS**]{.underline}

To understand the data in a better way , I created the following table to explain the description,data type, and the value of each feature.

+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| [*Feature*]{.underline}                                  | [*Definition*]{.underline}                                         | [*Type*]{.underline} |
+==========================================================+====================================================================+======================+
| **Age**                                                  | Age of subject                                                     | Numerical            |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Sex**                                                  | Gender of subject:\                                                | Nominal              |
|                                                          | 0 = female 1 = male                                                |                      |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Chest-pain type**                                      | Type of chest-pain experienced by the individual:\                 | Nominal              |
|                                                          | 1 = typical angina\                                                |                      |
|                                                          | 2 = atypical angina\                                               |                      |
|                                                          | 3 = non-angina pain\                                               |                      |
|                                                          | 4 = asymptomatic angina                                            |                      |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Resting Blood Pressure**                               | Resting blood pressure in mm Hg                                    | Numerical            |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Serum Cholesterol**                                    | Serum cholesterol in mg/dl                                         | Numerical            |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Fasting Blood Sugar**                                  | Fasting blood sugar level relative to 120 mg/dl:                   | Nominal              |
|                                                          |                                                                    |                      |
|                                                          | 0 = fasting blood sugar \<= 120 mg/dl\                             |                      |
|                                                          | 1 = fasting blood sugar \> 120 mg/dl                               |                      |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Resting ECG**                                          | Resting electrocardiographic results\                              | Nominal              |
|                                                          | 0 = normal\                                                        |                      |
|                                                          | 1 = ST-T wave abnormality\                                         |                      |
|                                                          | 2 = left ventricle hyperthrophy                                    |                      |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Max Heart Rate Achieved**                              | Max heart rate of subject                                          | Numerical            |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Exercise Induced Angina**                              | 0 = no 1 = yes                                                     | Nominal              |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **ST Depression Induced by Exercise Relative to Rest**   | 1.  ST Depression of subject                                       | Numerical            |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Peak Exercise ST Segment**                             | 1 = Up-sloaping\                                                   | Nominal              |
|                                                          | 2 = Flat\                                                          |                      |
|                                                          | 3 = Down-sloaping                                                  |                      |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Number of Major Vessels (0-3) Visible on Flouroscopy** | Number of visible vessels under flouro                             | Nominal              |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Thal**                                                 | Form of thalassemia:Â \                                             | Nominal              |
|                                                          | 3 = normal\                                                        |                      |
|                                                          | 6 = fixed defect\                                                  |                      |
|                                                          | 7 = reversible defect                                              |                      |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+
| **Diagnosis of Heart Disease**                           | Indicates whether subject is suffering from heart disease or not:\ | Nominal              |
|                                                          | 0 = absence\                                                       |                      |
|                                                          | 1, 2, 3, 4 = heart disease present                                 |                      |
+----------------------------------------------------------+--------------------------------------------------------------------+----------------------+

-   [**TYPE OF STUDY**]{.underline}

***This is observational study.***

-   [**DEPENDENT VARIABLE**]{.underline}

**Diagnosis of Heart Disease *is the output response (dependent ) variable and it is qualitative.***

-   [**INDEPENDENT VARIABLE**]{.underline}

***The independent variable is age ,gender,and all other variables.***

**CHECK IF THERE ARE MISSING VALUES (NA) IN THE <DATA:->**

```{r}
sum(is.na(data))
```

***Hence,there is no missing values in this data.***

**COUNT THE NUMBER OF PATIENTS HAVE OR HAVE NOT BEEN DIAGNOSED WITH HEART DISEASE:-**

```{r}
data%>%count(Diagnosis_Heart_Disease)
```

***Hence,164 patients do not have diagnosed heart disease and 139 patients have diagnosed with heart disease.***

**THE PROPORTION OF PATIENTS WITH CHEST PAIN TYPES:-**

```{r}
data%>%group_by(Chest_Pain_Type)%>%summarise(percent=100*n()/nrow(data))

```

***Hence, there are 8% of patients with typical angina chest pain, 16% of patients with atypical angina , 28% with non-angina chest pain and 48% with asymptomatic angina chest pain.***

**THE PROPORTION OF FEMALE AND MALE PATIENTS IN THE DATASET:-**

```{r}
data%>%group_by(Sex)%>%summarise(percent=100*n()/nrow(data))
```

***Hence ,there are 32 % females and 68 % males in the dataset.***

**CHECK THE PERCENTAGE OF MALES AND FEMALES WITH HEART DISEASE:-**

```{r}
sub_female=table(data[data$Sex==0,]$Diagnosis_Heart_Disease)
sub_female
sub_male=table(data[data$Sex==1,]$Diagnosis_Heart_Disease)
sub_male
H=c(25,114)
D=c(72,92)
c=data.frame(H,D)
colnames(c)=c("Have Heart Disease","Does not have Heart Disease")
rownames(c)=c("Females","Males")
c

```

***There are 25 females out of 97 who have diagnosed with heart disease and 114 males out of 206 were diagnosed with heart disease.***

***This indicates that 55.3 % of males in this dataset are diagnosed with heart disease where is only 26% of females are diagnosed with heart disease.***

-   [**FINDING 1:**]{.underline}

```         
**Males are more diagnosed with heart disease than females.**
```

***The summary function displays useful information about each feature such as: the minimum value, maximum value,first and third quartile, mean and the median.***

```{r}
summary(data)
```

***The describe function displays useful information about each feature such as: the minimum value, maximum value,standard deviation, number of observations(this is an easy tool to check if there is missing values),mean and the median and other useful information.***

```{r}
describe(data)
```

-   **Find the probability of a patient to have a heart disease \<\< 50**

```{r}
summary(data$Age)
```

```{r}
sd(data$Age)
```

-   ***(Âµ= 54.44 ,ðž¼=9.03)***

-   ***In order to see the probability of a patient to have a heart disease \<\< 50, we need to calculate Z=(x-Âµ)/ðž¼ .***

```{r}
pnorm(50,54.44,9.038662)

```

-   [**DATA VISUALIZATION**]{.underline}

***This is a barplot,helps to visualize the distribution of heart disease diagnosis:***

```{r}
data$Diagnosis_Heart_Disease[data$Diagnosis_Heart_Disease>0]=1
barplot(table(data$Diagnosis_Heart_Disease),main="Heart disease distribution",col="blue")
```

***This is a mosaic plot, helps to visualize the statistical association between two variables:***

```{r}
mosaicplot(data$Sex~data$Diagnosis_Heart_Disease,main="Heart disease outcome by gender",shade=F,color=T,xlab="Gender",ylab="Heart disease")
```

***This is a boxplot to display the age distribution of heart diagnosis:***

```{r}
boxplot(data$Age~data$Diagnosis_Heart_Disease,main="Heart disease distribution by Age",xlab="Heart disease Diagnosed",ylab="Age")
```

***This plot is to visualize the Heart Disease diagnosis Distributions by Chest pain. There are four types of chest pain:***

***1 = typical angina\
2 = atypical angina\
3 = non-angina pain\
4 = asymptomatic angina***

```{r}
data$Sex=as.factor(data$Sex)
data$Diagnosis_Heart_Disease=as.factor(data$Diagnosis_Heart_Disease)
data$Chest_Pain_Type=as.factor(data$Chest_Pain_Type)
ggplot(data,aes(x=Chest_Pain_Type,fill=Chest_Pain_Type))+geom_bar(position="fill")+labs(title="Heart disease diagnosis distributions by Chest Pain ",x="Heart disease Diagnosis",y="Chest Pain")+theme_test()
```

***Another plot to visualize heart disease diagnosis distributions by Number of major vessels:***

```{r}
data$Num_Major_Vessels_Flouro=as.factor(data$Num_Major_Vessels_Flouro)
ggplot(data,aes(x=Diagnosis_Heart_Disease,fill=Num_Major_Vessels_Flouro))+geom_bar(position="fill")+labs(title="Heart disease diagnosis distributions by Number of major vessels ",x="Heart disease Diagnosis",y="thal")+theme_test()
```

***Histogram of patient's age and gender:***

```{r}
attach(data)
data$Sex=as.factor(data$Sex)
#making a new dataframe to store the mean ages of the male and female patients so that it can be included in the ggplot face-wrap function
meanAge=data.frame(sex=c(0,1),age=c(mean(data[data$Sex==0,]$Age,mean(data[data$Sex==1,]$Age))))
meanAge
#ggplot of age of the patients categorized by sex
p=ggplot(data,aes(x=Age,fill=as.factor(Sex)))+geom_histogram(alpha=0.5,position="identity")+facet_wrap(~as.factor(Sex))+labs(title="Histogram of patient's age by gender",x="Age of patient's",y="Count",fill="Sex")+scale_fill_discrete(breaks=c("0","1"),labels=c("0-female","1-Male"))

p
```

***Heart disease diagnosis frequency by Resting electrocardiographic results and sex***

```{r}

data$Resting_ECG=as.factor(data$Resting_ECG)
data%>%ggplot(aes(x=Diagnosis_Heart_Disease,fill=Resting_ECG))+geom_bar(position="dodge")+facet_grid(~Sex)+scale_fill_brewer(palette="Dark2")+labs(title="Heart disease diagnosis frequency by resting ecg and Sex")
```

-   [**Plot for Correlations**]{.underline}

```{r}
data$Age=as.numeric(data$Age)
data$Sex=as.numeric(data$Sex)
data$Chest_Pain_Type=as.numeric(data$Chest_Pain_Type)
data$Resting_Blood_Pressure=as.numeric(data$Resting_Blood_Pressure)
data$Serum_Cholesterol=as.numeric(data$Serum_Cholesterol)
data$Fasting_Blood_Sugar=as.numeric(data$Fasting_Blood_Sugar)
data$Resting_ECG=as.numeric(data$Resting_ECG)
data$Max_Heart_Rate_Achieved=as.numeric(data$Max_Heart_Rate_Achieved)
data$Exercise_Induced_Angina=as.numeric(data$Exercise_Induced_Angina)
data$ST_Depression_Exercise=as.numeric(data$ST_Depression_Exercise)
data$Peak_Exercise_ST_Segment=as.numeric(data$Peak_Exercise_ST_Segment)
data$Num_Major_Vessels_Flouro=as.numeric(data$Num_Major_Vessels_Flouro)
data$Thalassemia=as.numeric(data$Thalassemia)
correlations=cor(data[,1:13])

```
